---
title: "TNT input file creation from R and processing of output for reading in R"
author: "Vera Weisbecker"
date: "06/05/2021"
output: html_document
---


```{r}

library(geomorph) # Geometric Morphometric Analyses of 2D/3D Landmark Data


#library(landvR)

# My rmd file for some reason won't recognise it s local file, so this sets the wd to local
library(rstudioapi) # Safely Access the RStudio API
setwd(dirname(getActiveDocumentContext()$path))

source("0_2_Utilities.r")

load (file = "../Data/Processed/processed_data_incl_allometry.rda")




writeland.tnt(GPA_Allmean$coords, file= "../Data/Results/Full_config_GPA.tnt")
writeland.tnt(GPA_Basimean$coords, file= "../Data/Results/Basi_config_GPA.tnt")
writeland.tnt(GPA_Restmean$coords, file= "../Data/Results/Rest_config_GPA.tnt")





```

#TNT outputs files that have two spaces befor a semicolon and this prevents read.nex to read them. Use the ReplaceInFiles addin every time before you attempt to read in the TNT output in R.

#FIRST you need to remove spaces before the semi-colons in teh output files using the ReplaceInFiles algorithm. First look for trees ; and replace with trees;, then for end ; and same. You can do this at a high level (all TNT files) and just stipulate the files endings you want affected (e.g. *with_NodeLabels.tree)

#Afterwards, the files need saving. It is better to do this separately so only the tree files get saved.


```{r setup, include=FALSE}

devtools::install_github("skranz/ReplaceInFiles")


```