---
title: "02_04 cophenetic tree comparisons"
author: "Vera Weisbecker"
date: "2022-12-08"
output: html_document
---

```{r}

library(ape)      # CRAN v5.6-2 
library(vegan)    # CRAN v2.6-2
library(geomorph) # [github::geomorphR/geomorph] v4.0.4 # CRAN v4.0.4 
library(TreeDist) # CRAN v2.5.0 

# My rmd file for some reason won't recognise it s local file, so this sets the wd to local
library(rstudioapi) # Safely Access the RStudio API
setwd(dirname(getActiveDocumentContext()$path))


source(file="0_2_Utilities.r")

load(file="../Data/Processed/Trees_collapsed.Rda")

```

#Here is the code for mantel test-based matrix correlation of distance matrices. Note that I checked and re-rooting the TNT trees isn't necessary, the results are the same. Note also that the cophenetic matrices of the non- NJ/UPGMA trees are not in the same order as the specimens in the morph datasets.

```{r}

Reference_cophen <- cophenetic(Full_tree)

Trees_for_cophen <- c( Mol_based_trees, 
                       TNT_full = list(reduced_TNT_trees[[1]]$GPA_Allmean), 
                       TNT_basi= list(reduced_TNT_trees[[1]]$GPA_basi),
                       TNT_Rest = list (reduced_TNT_trees[[1]]$GPA_Restmean),
                       NJ_full = list(reduced_morph_clusters[[1]][[2]]$GPA_Allmean),
                       NJ_basi = list(reduced_morph_clusters[[1]][[2]]$GPA_basi),
                       NJ_Rest = list(reduced_morph_clusters[[1]][[2]]$GPA_Restmean),
                       UPGMA_full = list(reduced_morph_clusters[[1]][[1]]$GPA_Allmean),
                       UPGMA_basi = list(reduced_morph_clusters[[1]][[1]]$GPA_basi),
                       UPGMA_Rest = list(reduced_morph_clusters[[1]][[1]]$GPA_Restmean)
                       
                       )

Trees_for_cophen_res <- c( Mol_based_trees, 
                       TNT_full = list(reduced_TNT_trees[[1]]$Allom_all_plus_consensus), 
                       TNT_basi= list(reduced_TNT_trees[[1]]$Allom_basi_plus_consensus),
                       TNT_Rest = list (reduced_TNT_trees[[1]]$Allom_rest_plus_consensus),
                       NJ_full = list(reduced_morph_clusters[[1]][[2]]$Allom_all_plus_consensus),
                       NJ_basi = list(reduced_morph_clusters[[1]][[2]]$Allom_basi_plus_consensus),
                       NJ_Rest = list(reduced_morph_clusters[[1]][[2]]$Allom_rest_plus_consensus),
                       UPGMA_full = list(reduced_morph_clusters[[1]][[1]]$Allom_all_plus_consensus),
                       UPGMA_basi = list(reduced_morph_clusters[[1]][[1]]$Allom_basi_plus_consensus),
                       UPGMA_Rest = list(reduced_morph_clusters[[1]][[1]]$Allom_rest_plus_consensus)
                       
                       )

```


# David Polly's suggestion for drilling into the detail of how the Morph clusters relate to their euclidean distances

```{r}

#"First calculate the pairwise morphometric distances between the taxa (A-B, A-C, B-C).  This distance is usually a Euclidean distance, and in geometric morphometrics that is the same as the pairwise Procrustes distances. "


Procdist_matrices <- vector("list", 6)

names(Procdist_matrices) <- names(Partitions)

for (i in 1:length(names(Partitions))){

    Procdist_matrices[[i]] <- as.matrix(dist(two.d.array( Partitions[[i]])))


}

#Check that colnames and rownames are in the same order throughout

rownames(Procdist_matrices$GPA_Allmean)==colnames(Procdist_matrices$GPA_Basimean==rownames(Procdist_matrices$GPA_Restmean==colnames(Procdist_matrices$Allom_all_plus_consensus)))

#"Then measure the patristic distances on one of the trees (i.e., the sum of the branch lengths connecting them) in the same order (A-B, A-C, B-C).  "
#this is just using the cophenetic trees from the analysis above, and we need to make sure that the matrix has the same order


#This makes it easier to run the code in one hit - assemble just the morphological trees from chunk 12
Trees_for_patristic <- append(Trees_for_cophen[-c(1:length(Mol_based_trees))],Trees_for_cophen_res[-c(1:length(Mol_based_trees))])


#Hard coding the numbers of trees here for distinguishing between just GPA and residuals
names(Trees_for_patristic) <- c(names(Trees_for_patristic[1:9]),paste(names(Trees_for_patristic[1:9]), "_res", sep=""))

#A little check to ensure the co-phenetic trees have been taken into the right position in their list
TreeDistance(Trees_for_patristic$UPGMA_Rest, Trees_for_cophen$UPGMA_Rest)
TreeDistance(Trees_for_patristic$UPGMA_full_res, Trees_for_cophen_res$UPGMA_full)


#Creating list of patristic (co-phenetic) matrices
Patristic_matrices <-  vector("list", length(Trees_for_patristic))

names(Patristic_matrices) <- names(Trees_for_patristic)

for (i in 1: length(names(Trees_for_patristic))){
  
  Patristic_matrices[[i]] <- cophenetic.phylo(Trees_for_patristic[[i]])
  
  
}


#But now the matrix columns are not all in order - they have to be for correct comparison with each other. This affects the TNT trees but not the UPGMA/NJ trees but running it all here

colnames(Procdist_matrices$GPA_Allmean)==colnames(Patristic_matrices$TNT_full)

#We need to standardise the matrix using match(). I'm jusing GPA_Allmean as standard because it reflects the order of all morphological data, but it doesn't matter as long as the matrices all have the same row/column names

for (i in 1: length(Trees_for_patristic)){
  
  Patristic_matrices[[i]] <- Patristic_matrices[[i]] [match(rownames(Procdist_matrices$GPA_Allmean), rownames(Patristic_matrices [[i]] )),
                                                    match(colnames(Procdist_matrices$GPA_Allmean), colnames(Patristic_matrices [[i]] ))]
  
}

#column names are now OK

colnames(Procdist_matrices$GPA_Allmean) == colnames(Patristic_matrices$TNT_full)
colnames(Procdist_matrices$GPA_Allmean) == colnames(Patristic_matrices$TNT_Rest)

# Also, individual cells have the correct values in them - this is just a random spot check where different cells. 

Matrix_for_checking <- Patristic_matrices$TNT_full

unordered_cophen_matrix <- cophenetic.phylo(Trees_for_patristic$TNT_full)

#If the below is TRUE, the distance values in the re-arranged patristic matrices are identical despite being in different rows/columns relative to the original, unordered cophenetic matrix

unordered_cophen_matrix[which(colnames(unordered_cophen_matrix)=="Macrotis_lagotis"), 
     which(rownames(unordered_cophen_matrix)=="Hypsiprymnodon_moschatus")]  ==
  
Matrix_for_checking[which(colnames(Matrix_for_checking)=="Macrotis_lagotis"),
                    which(rownames(Matrix_for_checking)=="Hypsiprymnodon_moschatus")]
  

remove(Matrix_for_checking, unordered_cophen_matrix)

```

#First set up a list of multiple Procdist matrices, of which there are only 6, so that they will run according to the 18 trees we're comparing against. First 9 non-size adjusted, then 9 size-adjusted trees

```{r}



Procdist_matrices <-list( Procdist_matrices[1:3], Procdist_matrices[1:3], Procdist_matrices[1:3], Procdist_matrices[4:6], Procdist_matrices[4:6], Procdist_matrices[4:6])

Procdist_matrices <- unlist(Procdist_matrices, recursive=FALSE)


#Making sure the order of Procdist matrices and Patristic matrices to be compared matches, just checking visually
rbind(names(Procdist_matrices), names(Patristic_matrices))

#The below runs the vector correlations, but this is just for double checking that the mantel test below works. Good to do it two ways I guess ;-)

#vector_correlations <- vector()
#
# for (i in 1:length(Procdist_matrices)){
#
# vector_correlations[i] <- cor(c(Procdist_matrices[[i]]), c( Patristic_matrices[[i]]) )
#
# }

# names(vector_correlations)<- names(Patristic_matrices)
#
# vector_correlations
# vector_correl_matrix <- matrix(vector_correlations, nrow=9)
# vector_correl_matrix <- as.data.frame(vector_correl_matrix)
# rownames(vector_correl_matrix)<- c("Parsimony-Full", "Parsimony-Basicr.", "Parsimony-Rest","NJ-Full", "NJ-Basicr.", "NJ-Rest","UPGMA-Full", "UPGMA-Basicr.", "UPGMA-Rest" )
# colnames(vector_correl_matrix) <- c("GPA", "Residuals")


#write.csv(vector_correl_matrix, file="../Results/Patristic_cophenetic_vector_correls.csv ")

```

#make cophenetic matrices for molecular-based vs reference (1-4) and procdist matrices:

```{r}

cophen_matrices <- list()

#cophen.phylo an cophen do the same thing

for (i in 1: length(Trees_for_cophen[1:4])){
  
 cophen_matrices[[i]] <- cophenetic.phylo(Trees_for_cophen[[i]])
  
}


names(cophen_matrices) <- names( Trees_for_cophen[1:4])

#The column names don't match up

colnames(Reference_cophen) == colnames(cophen_matrices$Molecular_TNT_raw)


for (i in 1: length(cophen_matrices)){
  
 cophen_matrices[[i]] <- cophen_matrices[[i]] [match(rownames(Reference_cophen), rownames(cophen_matrices [[i]] )),
                                                    match(colnames(Reference_cophen), colnames(cophen_matrices [[i]] ))]
  
}

#Now the columns and rows are correct
colnames(Reference_cophen) == colnames(cophen_matrices$Molecular_TNT_raw)
rownames(Reference_cophen) == rownames(cophen_matrices$Molecular_TNT_raw)


Mantel_temp <- list()
Mantel_corrs <- matrix(nrow=4, ncol=2)
rownames(Mantel_corrs)<- names(cophen_matrices) 
colnames(Mantel_corrs)<- c("Mantel statistic", "p")

for (i in (1: length (cophen_matrices))){
  
  Mantel_temp<- mantel(Reference_cophen,cophen_matrices[[i]], permutations=10000 )
  
  Mantel_corrs[i,] <- c(Mantel_temp$statistic, Mantel_temp$signif)
  
}







rownames(Mantel_corrs)<- c("Molecular, Parsimony", "Molecular, Raw Distances", "Molecular, GTR", "Molecular, GTRig")
Mantel_corrs

write.csv(Mantel_corrs, file= "../Results/Matrix_correls_Mol_Ref.csv")

```



#Mantel test of distance matrices. Note that the first matrices in the Mantel tests are the ones being permuted (i.e. they are the reference matrices)

```{r}

#Starting with patristic versus procrustes distance

Mantel_correls_patr_procd <- list()

for (i in 1:length(Patristic_matrices)){
  
 Mantel_correls_patr_procd [[i]] <- mantel(Patristic_matrices[[i]], Procdist_matrices[[i]], permutations = 10000)
  
  
}


#"As a point of comparison, one could also calculate cophenetic correlation between pairwise morphometric distances and molecular phylogeny, which would be a measure of how closely the morphometric structure matches the “real” phylogeny."


#First check that the column names of the reference cophenetic matrix and the patristic matrices are matching

colnames(Patristic_matrices$TNT_full)==colnames(Reference_cophen)

Mantel_reference_patr <- list()

for (i in 1: length(Patristic_matrices)){
  
  Mantel_reference_patr [[i]] <- mantel(Reference_cophen, Patristic_matrices[[i]],  permutations = 10000)
  
  
}

#I am here also adding the procrustes distances vs reference phylogeny to have it all in one table. This repeats the same analyses three times but is easier to tabulate below.

#again, just checking

colnames(Procdist_matrices$GPA_Allmean)==colnames(Reference_cophen)

Mantel_reference_procd <- list()

for (i in 1: length(Procdist_matrices)){
  
  Mantel_reference_procd [[i]] <- mantel(Reference_cophen, Procdist_matrices[[i]],  permutations = 10000)
  
  
}

```


#Now make table. A lot of work but you never have to do it again.

```{r}

#First, set up matrix

matrix_Mantel <- matrix(ncol = 13, nrow=9)

#enter the patristic vs proc distance tests

for (i in 1:(length(Mantel_correls_patr_procd)/2)){
  
  matrix_Mantel[i,1:2] <- c(Mantel_correls_patr_procd[[i]]$statistic, Mantel_correls_patr_procd[[i]]$signif)
  
}

for (i in 1:(length(Mantel_correls_patr_procd)/2)){
  
  matrix_Mantel[i,3:4] <- c(Mantel_correls_patr_procd[[i+9]]$statistic, Mantel_correls_patr_procd[[i+9]]$signif)
  
}

#enter the patristic vs. reference


for (i in 1:(length(Mantel_correls_patr_procd)/2)){
  
  matrix_Mantel[i,5:6] <- c(Mantel_reference_patr[[i]]$statistic, Mantel_reference_patr[[i]]$signif)
  
}

for (i in 1:(length(Mantel_correls_patr_procd)/2)){
  
  matrix_Mantel[i,7:8] <- c(Mantel_reference_patr[[i+9]]$statistic, Mantel_reference_patr[[i+9]]$signif)
  
}


#enter procrustes distances vs reference, leaving one column blank because the last column is a bit different

for (i in 1:3){
  
  matrix_Mantel[i,10:11] <- c(Mantel_reference_procd[[i]]$statistic, Mantel_reference_procd[[i]]$signif)
  
}

for (i in 1:3){
  
  matrix_Mantel[i,12:13] <- c(Mantel_reference_procd[[i+9]]$statistic, Mantel_reference_procd[[i+9]]$signif)
  
}


#Actually all significances are at 0.0000 so I can save a column

matrix_Mantel <- matrix_Mantel[,-c(2,4,6,8,11,13)]

#And we can round all the numbers to 2 decimals
matrix_Mantel <-round(matrix_Mantel, digits=2)

#Also add in the molecular comparison and label these
matrix_Mantel[4:7,6] <- Mantel_corrs[,1]
  
#To allow naming everything, take to data frame

matrix_Mantel <- as.data.frame(matrix_Mantel)

rownames(matrix_Mantel)<- c("Parsimony-Full", "Parsimony-Basicr.", "Parsimony-Rest","NJ-Full", "NJ-Basicr.", "NJ-Rest","UPGMA-Full", "UPGMA-Basicr.", "UPGMA-Rest" )
colnames(matrix_Mantel) <- c("Patristic vs Procrustes distances", "","Patristic vs molecular reference", "","", "Procrustes distances/Mol. distances vs molecular reference")
#Label the last column differently
matrix_Mantel[,5]<- c("Full config.","Basicranium","Rest","Molecular, Parsimony", "Molecular, Raw Distances", "Molecular, GTR", "Molecular, GTRig", "","")



matrix_Mantel

matrix_Mantel <- rbind (c("Statistic - GPA", "Statistic- Residuals","Statistic - GPA", "Statistic- Residuals","","Statistic - GPA", "Statistic- Residuals"), matrix_Mantel)

rownames(matrix_Mantel)[1]<-""
#looking good?

matrix_Mantel

#save it!

write.csv(matrix_Mantel, file="../Results/Patristic_cophenetic_mantel_correls.csv ")

```


Again, this code requires no saving 
